---
title: "Bayesian Models"
output:
  html_document:
    df_print: paged
date: "2023-08-04"
editor_options:
  chunk_output_type: console
---

## Data Read In

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source(here::here("Helper_Functions/data-readin.R"))
source(here::here("Helper_Functions/analysis-functions.R"))
library(tidyverse)
library(ggplot2)
library(CARBayes)
library(viridis)
library(cowplot)
library(coda)

```

```{r data_readin, include=FALSE}

# DP05_0065PE Percent!!Race alone or in combination with one or more other races!!Total population!!Black or African American

ga_map_data <- data_readin(acs_vars = "DP03_0119PE,DP04_0003PE,DP03_0009PE,DP05_0037PE,DP05_0038PE,DP05_0070PE,DP05_0071PE,DP05_0065PE", acs_yr = 2020, mort_year = 2020)

```

## Map Components

```{r weight matrix}
# Bayes---------------------------------------------------

ga_map_nb <- spdep::poly2nb(ga_map_data, queen = F)

W_mt <- spdep::nb2mat(ga_map_nb, style = "B")

# Print the smoothed map
ps <- ggplot(ga_map_data) + 
  geom_sf(aes(fill = incidence * 1000)) +
  labs(title = "Unsmoothed Smoothed\n Mortality Rates 19")

# plot(ga_map_nb)

print(ps)

# ga_map_data %>% arrange(desc(incidence)) %>% slice(1:20) %>% mutate(od_rank = row_number())

```


## Simpler Model

### Frequentist Model

```{r vacancy_model}

form <- "mortality ~ vacancy_rate_std + offset(log(population))"

model <- glm(as.formula(form), family = "poisson",
            data = ga_map_data)

print(summary(model))

```


### Bayesian Model


```{r bayesmodel1term}

form <- "mortality ~ vacancy_rate_std + offset(log(population))"

run_model <- function(form){
  model.spatial <- S.CARleroux(formula = as.formula(form), family = "poisson",
                               data = ga_map_data,
                               W = W_mt, burnin = 0,
                               n.sample = 1000000,
                               thin = 5)
  
  return(model.spatial)
  }

iter_1 <- run_model(form)
iter_2 <- run_model(form)

print(iter_1)
iter_1$summary.results

# https://stats.stackexchange.com/questions/266749/mcmc-convergence

```


```{r bayeschainanalysis}

samples_beta <- mcmc.list(iter_1$samples$beta, iter_2$samples$beta)

plot(samples_beta[,2], density = F, trace = F)
gelman.diag(samples_beta[,2])

```

## Full Model

### Frequentist


```{r freq_model}

form <- "mortality ~ pct_poverty_std + vacancy_rate_std + unemployment_rate_std + offset(log(population))"

model <- glm(as.formula(form), family = "poisson",
            data = ga_map_data)

print(summary(model))

```


According to the frequentist model, both percent in poverty and vacancy rate were statistically associated with the rate of opioid overdoses.

### Bayesian

Note that the S.CARleroux function currently produces a warning: `Warning message: In mat2listw(W) : style is M (missing); style should be set to a valid value`. But you can ignore it, because it is produced by `mat2listw(W)` not being given a style argument in `getAnywhere(poisson.lerouxCAR)`. But I have checked and a square matrix goes into that function and it's 

```{r model}

form <- "mortality ~ pct_poverty_std + vacancy_rate_std + unemployment_rate_std + offset(log(population))"

run_model <- function(form){
  model.spatial <- S.CARleroux(formula = as.formula(form), family = "poisson",
                               data = ga_map_data,
                               W = W_mt, burnin = 20000,
                               n.sample = 1e6,
                               thin = 5)
  
  return(model.spatial)
  }

iter_1 <- run_model(form)
iter_2 <- run_model(form)

print(iter_1)

iter_1$summary.results


```


```{r modeldiag}

samples_beta <- mcmc.list(iter_1$samples$beta, iter_2$samples$beta)

plot(samples_beta[,2:4], density = FALSE)
gelman.diag(samples_beta[,2:4])

```

### Plot from Bayesian

```{r}

# Georgia Map Data fitted vs actual
ga_map_data_res <- ga_map_data %>% 
   mutate(mortality_hat = fitted(iter_1)) %>% 
  relocate(mortality_hat, .after = mortality) %>% 
  mutate(across(c(mortality, mortality_hat), ~ .x / population * 100000, .names = "inc_{.col}"))

ga_map_data_res_long <- ga_map_data_res %>% 
  st_drop_geometry() %>% 
  dplyr::select(NAME, rucc_code13, mortality, mortality_hat, inc_mortality, inc_mortality_hat, population, pct_poverty_std:unemployment_rate_std) %>% 
  pivot_longer(cols = c(inc_mortality, inc_mortality_hat), names_to = "Response")

ggplot(ga_map_data_res_long, aes(x = NAME, y = value, group = Response, colour = Response)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))

```



```{r}

plot_maps(ga_map_data_res, iter_1, mortality, mortality_pop)

```





## Model with RUCC Codes

```{r urcccodes, echo=FALSE}

# Same with rucc codes
ggplot(ga_map_data) +
  geom_sf(aes(fill = rucc_code13)) +
  scale_fill_viridis(discrete = T, direction = -1) +
  labs(title = "NCHS Urban Rural Classification")

ga_map_data %>% 
  st_drop_geometry() %>% 
  count(rucc_code13) %>% mutate(prop = n / sum(n) * 100)

ga_map_data %>% 
  st_drop_geometry() %>% 
  group_by(rucc_code13) %>% 
  ggplot(aes(x = rucc_code13, y = incidence)) +
  geom_boxplot() +
  labs(title = "Opioid Mortality Incidence by Urban Rural",
       subtitle = "Large Central, Large Fringe, Medium Metro, Small Metro, Micrro, Non-Core",
       x = "URCC Code",
       y = "Incidence")
  


```

```{r bayesrucc}

form <- "mortality ~ rucc_code13 + pct_poverty + vacancy_rate + unemployment_rate+ offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data,
                             W = W_mt, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data,
                             W = W_mt, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
# plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])


model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(model)

model.spatial$summary.results

```


```{r map}
ga_map_data <- mutate(ga_map_data,
                      fitted_values = model.spatial$fitted.values)

ggplot(ga_map_data, aes(x = mortality, y = fitted_values)) +
  geom_point() +
  theme_bw()

test_vals_wide <- tibble(id = 1:length(model.spatial$fitted.values), 
                    fitted_inc = model.spatial$fitted.values,
                    true_inc = ga_map_data$mortality)

test_vals <- test_vals_wide %>% 
  pivot_longer(c(fitted_inc, true_inc))

ggplot(test_vals, aes(x = id, y = value, group = name, colour = name)) +
  geom_point() +
  theme_bw()

p1 <- ggplot(ga_map_data) +
  geom_sf(aes(fill = mortality / population)) +
  scale_fill_viridis()

p2 <- ggplot(ga_map_data) +
  geom_sf(aes(fill = fitted_values / population)) +
  scale_fill_viridis() 

plot_grid(p1, p2, labels = c("Unsmoothed", "Smoothed"))

```

```{r variancereduction}

var.test(test_vals$true_inc, test_vals_wide$fitted_inc, alternative = "two.sided")

```

# Distance Metrics Added

## Map Components

```{r disttoroad}

ga_map_data_short <- ga_map_data %>% 
  filter(!is.na(dist_to_usroad))

# The following two counties were removed from the adjacency matrix.
# ga_map_data %>% 
#   st_drop_geometry() %>% 
#   mutate(county_id = row_number()) %>% 
#   dplyr::select(county_id, COUNTY, dist_to_usroad) %>% 
#   filter(is.na(dist_to_usroad))
# county_id   COUNTY dist_to_usroad
# 1       158 Columbia             NA
# 2       159   Echols             NA 

```


```{r weightmatrixdist}
# Bayes---------------------------------------------------

ga_map_nb_short <- spdep::poly2nb(ga_map_data_short, queen = F)

W_mt_short <- spdep::nb2mat(ga_map_nb_short, style = "B")

# Print the smoothed map
# ps <- ggplot(ga_map_data) + 
#   geom_sf(aes(fill = incidence * 1000)) +
#   labs(title = "Unsmoothed\n Mortality Rates 19")

# plot(ga_map_nb)

print(ps)

# ga_map_data %>% arrange(desc(incidence)) %>% slice(1:20) %>% mutate(od_rank = row_number())

```



## Model with RUCC Codes and Distance

```{r distplots, echo=FALSE}

# Same with rucc codes
ggplot(ga_map_data_short) +
  geom_sf(aes(fill = rucc_code13)) +
  scale_fill_viridis(discrete = T, direction = -1) +
  labs(title = "NCHS Urban Rural Classification")

ggplot(ga_map_data_short) +
  geom_sf(aes(fill = dist_to_usroad)) +
  scale_fill_viridis(direction = -1) +
  labs(title = "US Survey Ft Distance to US Road")

ga_map_data %>% 
  st_drop_geometry() %>% 
  count(rucc_code13) %>% mutate(prop = n / sum(n) * 100)

ga_map_data %>% 
  st_drop_geometry() %>% 
  ggplot(aes(x = dist_to_usroad_std, y = incidence)) +
  geom_point() +
  labs(title = "Opioid Mortality Incidence by Urban Rural",
       subtitle = "Large Central, Large Fringe, Medium Metro, Small Metro, Micrro, Non-Core",
       x = "URCC Code",
       y = "Incidence")
  
```

## Dist to Road

```{r bayesdist-1}

form <- "mortality ~ dist_to_usroad_std + pct_poverty_std + vacancy_rate_std + unemployment_rate_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
# plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])


model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results

```

## Dist to Treatment (FINAL MODEL)

```{r bayesdist-2}

form <- "mortality ~ vacancy_rate_std + dist_to_treatment_std + pct_poverty_std + unemployment_rate_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10,
                             rho = .8)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10,
                             rho = .8)

# Need to add comparison chain
# plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])


model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

exp(model.spatial$summary.results[,1:3] *5)

```



# Based on Bivariate Analysis

```{r bayesdisttreat}

form <- "mortality ~  dist_to_treatment_std * rucc_code13_5 +  dist_to_usroad_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])

# plot(mcmc.list(model.spatial.chain2$samples$beta)[,-1])

model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results
model.spatial.chain2$summary.results

```


```{r bayesvacancyrate}

form <- "mortality ~  vacancy_rate_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])

# plot(mcmc.list(model.spatial.chain2$samples$beta)[,-1])

model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results
model.spatial.chain2$summary.results

```



```{r bayesruccsolo}

form <- "mortality ~  rucc_code13_5 + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])

# plot(mcmc.list(model.spatial.chain2$samples$beta)[,-1])

model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results
model.spatial.chain2$summary.results

```

```{r bayesvacancyrateint}

form <- "mortality ~  vacancy_rate_std * rucc_code13_5 +  dist_to_usroad_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             thin = 10)

# Need to add comparison chain
plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])

# plot(mcmc.list(model.spatial.chain2$samples$beta)[,-1])

model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results
model.spatial.chain2$summary.results

```


```{r bayesvacancyrate8}

form <- "mortality ~  vacancy_rate_std * rucc_code13_5 +  dist_to_usroad_std + offset(log(population))"

model.spatial <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             rho = .8,
                             n.sample = 1e6,
                             thin = 10)

model.spatial.chain2 <- S.CARleroux(as.formula(form), family = "poisson",
                             data = ga_map_data_short,
                             W = W_mt_short, burnin = 2000,
                             n.sample = 1e6,
                             rho = .8,
                             thin = 10)

# Need to add comparison chain
plot(mcmc.list(model.spatial$samples$beta, model.spatial.chain2$samples$beta)[,-1])

# plot(mcmc.list(model.spatial.chain2$samples$beta)[,-1])

model <- lm(as.formula(form), family = "gaussian",
            data = ga_map_data)

print(summary(model.spatial))
print(model.spatial)
print(summary(model))

model.spatial$summary.results
model.spatial.chain2$summary.results

```

```{r map short}

ga_map_data_short <- mutate(ga_map_data_short,
                      fitted_values = model.spatial$fitted.values)

test_vals_wide <- tibble(id = 1:length(model.spatial$fitted.values), 
                    fitted_inc = model.spatial$fitted.values,
                    true_inc = ga_map_data_short$mortality)

test_vals <- test_vals_wide %>% 
  pivot_longer(c(fitted_inc, true_inc))

ggplot(test_vals, aes(x = id, y = value, group = name, colour = name)) +
  geom_point() +
  theme_bw()

p1 <- ggplot(ga_map_data_short) +
  geom_sf(aes(fill = mortality / population)) +
  scale_fill_viridis()

p2 <- ggplot(ga_map_data_short) +
  geom_sf(aes(fill = fitted_values / population)) +
  scale_fill_viridis() 

plot_grid(p1, p2, labels = c("Unsmoothed", "Smoothed"))

```


```{r variancereduction short}

var(test_vals_wide$true_inc)

var(test_vals_wide$fitted_inc)

var.test(test_vals_wide$true_inc, test_vals_wide$fitted_inc, alternative = "two.sided")

```

